UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-06-26 ms/main.tex
some more FEV1 stuff plus distance from London

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Most admixture lines have coordinates in the south east of the UK, where the city of London is located, likely reflecting its high migrant population. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.

@@

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$). Distributions of FEV1 from Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.

@@

To reduce the potential risks for re-identification from results in this publication, data has been randomly permuted so that the population characteristics are preserved but individual-level data is not presented directly in the figures. We rounded each attribute to an attribute-specific number of bins, and then permuted the data in the following way: 
For each point (i.e. each individual) in UMAP visualizations, and each attribute, we identified the 9 nearest neighbouring points, and copied the attribute from a randomly selected neighbor (thus allowing for the possibility of one value being printed twice).  Because this process is done independently for each visualization, a given point shown on the figure will copy values from different randomly selected individuals. Additionally, spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km. For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}

@@

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/east_asia_ridgeplots_fev.png}
    \end{subfigure}
    \caption{Plots of the distributions of FEV1 by sex for East Asian populations. Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from figure~\ref{fig:supp_comparison_fev_chi_eur}. The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive. Asterisks indicate significant difference from the Japanese population, using Welch's unpaired t-test with a Bonferroni correction for 3 groups. The dashed lines are the means of the distributions and Japanese populations have consistently higher means, though the test between Chinese and Japanese males fell short of the significance threshold.}
    \label{fig:supp_fev_ridgeplots}
\end{figure*}

>>>>

Figure~\ref{fig:umap_ukbb_geo_osgb} presents the UMAP projection from figure~\ref{fig:umap_ukbb_pc10} coloured instead by geographical coordinates from the Ordnance Survey National Grid (OSGB1936), with distances defined as a north or east position relative to the Isles of Scilly. Geographic clusters form in the large White British grouping, reflecting the relationship between genetic and geographic distance, as has been observed in Europe-wide data\cite{novembre2008europe}. Colouring the projection by distance from London reflects the high migration levels to the city and surrounding area; individuals with more admixed backgrounds tend to have coordinates in the south east of the UK, as visible in figure~\ref{fig:supp_london_distance}. The detailed shape of extended clusters is not stable as we vary the number of PCs included, but the patterns mentioned above are preserved. Figures~\ref{fig:supp_montage_ukbb_eth}, \ref{fig:supp_montage_ukbb_ns}, and~\ref{fig:supp_montage_ukbb_ew} show UMAP plots using the top $40$ PCs from the UKBB.

@@

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though the sample of males is much smaller. Distributions of FEV1 by sex from selected Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.

@@

To reduce the potential risks for re-identification from results in this publication, data has been randomly permuted so that the population characteristics are preserved but individual-level data is not presented directly in the figures. We rounded each attribute to an attribute-specific number of bins, and then permuted the data in the following way: 
For each point (i.e. each individual) in UMAP visualizations, and each attribute, we identified the 9 nearest neighbouring points, and copied the attribute from a randomly selected neighbor (thus allowing for the possibility of one value being printed twice).  Because this process is done independently for each visualization, a given point shown on the figure will copy values from different randomly selected individuals. Projections coloured by participants' spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km. Projections coloured by participants' distance from London have random noise added (normally distributed about 0 with a standard deviation of 5km) before binning to the nearest km. For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.9\textwidth}
    \includegraphics[width=0.9\textwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_london_permuted_10nn_sd10_base1_maxdist200_2019625172227_resize.jpeg}
    \end{subfigure}
    \caption{UMAP on UKBB data, coloured by distance from London, with red representing those living closer to London and blue representing those living farther from London. A 200km radius extends roughly to Cardiff, and a 100km radius extends roughly to cities such as Leicester and Bath, and contains cities such as Oxford, Cambridge, and Peterborough. Data has been randomized as explained in the materials and methods section.}
    \label{fig:supp_london_distance}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.5\textwidth}
    \includegraphics[width=0.5\textwidth]{images/east_asia_population_colours.png}
    \end{subfigure}
    \caption{Individuals from the zoomed in section in figure~\ref{fig:supp_comparison_fev_chi_eur} used in statistical testing, coloured the same as in figure~\ref{fig:supp_fev_ridgeplots}. Brown, blue, and green represent those born in the Philippines, Malaysia, and Japan; pink represents those who self-identify as Chinese. The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive.}
    \label{fig:supp_fev_test_pops}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/east_asia_ridgeplots_fev.png}
    \end{subfigure}
    \caption{Plots of the distributions of residual FEV1 by sex for East Asian populations, after adjusting for height, age, age$^2$, and sex through linear regression. Individuals were limited to those in the "Chinese/Other Ethnic Group" cluster from figure~\ref{fig:supp_comparison_fev_chi_eur}. The Chinese individuals were those who self-identified their ethnic background as Chinese, and the remaining populations were determined based on country of birth; the categorizations are mutually exclusive. Asterisks indicate significant difference from the Japanese population, using Welch's unpaired t-test with a Bonferroni correction for 3 groups. The dashed lines are the means of the distributions, and Japanese populations have consistently higher means.}
    \label{fig:supp_fev_ridgeplots}
\end{figure*}
